\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-08 paper/ms.tex
words about performance

We have derived a novel method for computing the Cholesky factorization of the
covariance matrices produced by \celeriteterm\ models.
In this section, we demonstrate the scalability and numerical performance of
this algorithm.

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\facility{Kepler}
\software{%
     \project{AstroPy} \citep{Astropy-Collaboration:2013},
     \project{corner.py} \citep{Foreman-Mackey:2016},
     \project{Eigen} \citep{Guennebaud:2010},
     \project{emcee} \citep{Foreman-Mackey:2013},
     \project{george} \citep{Ambikasaran:2016},
     \project{LAPACK} \citep{Anderson:1999},
     \project{matplotlib} \citep{Hunter:2007},
     \project{numpy} \citep{Van-Der-Walt:2011},
     \project{transit} \citep{Foreman-Mackey:2016a},
     \project{scipy} \citep{Jones:2001}.
}

>>>>

We have derived a novel and scalable method for computing the Cholesky
factorization of the covariance matrices produced by \celeriteterm\ models.
In this section, we demonstrate the numerical and computational performance of
this algorithm.

First, we empirically confirm that the method solves systems of the form
$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
The left panel of \Figure{error} shows the L-infinity norm for the residual
$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
terms $J$.
For each pair of $N$ and $J$ we repeated the experiment 10 times with
different randomly generated datasets and \celeriteterm\ parameters and
averaged across experiments to remove some sampling noise from this figure.
The numerical error introduced by our method increases weakly for larger
datasets $\sim N^{0.15}$ and roughly linearly with $J$.

For the experiments with small $N \le 2048$, we also compared the log
determinant calculation to the result calculated using the general log
determinant function provided by the \project{NumPy} project
\citep{Van-Der-Walt:2011}, itself built on the \project{LAPACK}
\citep{Anderson:1999} implementation in the Intel Math Kernel
Library\footnote{\url{https://software.intel.com/en-us/intel-mkl}}.
The right-hand panel of \Figure{error} shows a histogram of the fractional
error in the log determinant calculation for each of the 480~experiments with
$N \le 2048$.
Across the board, these results agree within machine precision and the error
only weakly increases with $N$ and $J$.

We continue by measuring the real-world computational efficiency of this
method as a function of $N$ and $J$.
This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
i5 processor, but find similar performance on other platforms.
The \celerite\ implementation used for this test is the Python interface with
a C++ back end, but there is little overhead introduced by Python so we find
similar performance using implementations in C++ and Julia
\citep{Bezanzon:2012}.
\Figure{benchmark} shows the computational cost of evaluating
\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
The empirical scaling as a function of data size matches the theoretical
scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
some overhead for smaller models.
For comparison, the computational cost of the same computation using the Intel
Math Kernel Library, a highly optimized general dense linear algebra package,
is shown as a dashed line.
\celerite\ is faster than the reference implementation for all but the
smallest datasets with many terms.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\facility{Kepler}
\software{%
     \project{AstroPy} \citep{Astropy-Collaboration:2013},
     \project{corner.py} \citep{Foreman-Mackey:2016},
     \project{Eigen} \citep{Guennebaud:2010},
     \project{emcee} \citep{Foreman-Mackey:2013},
     \project{george} \citep{Ambikasaran:2016},
     \project{Julia} \citep{Bezanzon:2012},
     \project{LAPACK} \citep{Anderson:1999},
     \project{matplotlib} \citep{Hunter:2007},
     \project{numpy} \citep{Van-Der-Walt:2011},
     \project{transit} \citep{Foreman-Mackey:2016a},
     \project{scipy} \citep{Jones:2001}.
}
